Cultural Transmission and Technological Transitions during the Late Paleolithic in Korea

### 2022-11-30 analysis/paper/paper.qmd
remove PCA

To compute correlations between the attributes [@bettinger1999point; @eerkens2008cultural; @garvey2018current], we used Principal Component Analysis (PCA). PCA is commonly used to detect correlations among multiple variables, and has the advantage of allowing us to investigate the relationship between attributes as well as the impact of each attribute on the overall shape of stemmed points. PCA is often used to reduce dimensionality, detect patterns and clusters in data, indicate relationships between variables as well as cases, and identify which variables most influence these patterns [@cascalheira2018use; @suarez2019life]. PCA is widely used for geometric morphometric analysis of artifacts to extract general patterns of shape variation [@selden2014advances; @suarez2019life; @leplongeon2020backed; @ivanovaite2020all].

@@

```{r}
#| label: fig-plot-cv-over-time
#| fig-cap: CV values for the second and third chronological phases in the Korean Late Paleolithic period. The points represent CV values for each attribute. The vertical lines indicate the range of confidence intervals.
#| fig-width:   4.5
#| fig-height:  4.5

@@

```{r}
#| label: fig-plot-cv-four-assemblage
#| fig-cap: CV values of four attributes from assemblages that more than 5 stemmed points. Among the four assemblages, SYG1_2 and SYG6_2 are from Phase 2 of the Korean Late Paleolithic chronology and SYG1_3 and YS_3 are from Phase 3.
#| fig-width:   4.5
#| fig-height:  4.5

@@

```{r}
#| label: fig-plot-cv-by-site
#| fig-cap: CV values per site that has more than five stemmed points. The points represent CV values for each attribute. The vertical lines are confidence intervals.
#| fig-width:   8.5
#| fig-height:  4.5

@@

cv_plot_site_label %>%
  ggplot(aes(x = variable,
             y = cv_by_site)) +
  geom_point() +
  geom_errorbar(aes(ymin = cv_low_sharma,
                    ymax = cv_high_sharma),
                width = 0.2) +
  xlab("Attribute") +
  ylab("Coefficient of Variation on Attributes") +
    # draw a line to show our threshold value
  geom_hline(yintercept = 25, colour = "red") +
  facet_wrap( ~ label, scales = "free_y") +
  theme(legend.position = "none") +
  coord_cartesian(ylim = c(0,100)) +
  theme_bw() +
  theme(strip.text.x = element_text(size = 6.5))
```

@@

We examined the relationship between CV values and raw materials to test the hypothesis that the shape of stemmed points was highly dependent on raw material. We excluded raw materials that were used for less than three points. @fig-plot-cv-by-raw-material shows that quartzite, the raw materials for stemmed points at Sachang, has the highest CV value with a wide range of conference intervals. Other raw materials are generally stable in variation throughout the attributes.

```{r}
#| label: fig-plot-cv-by-raw-material
#| fig-cap: CV values for raw materials that were used for making more than one stemmed point. The points represent CV values for each attribute. The vertical lines indicate the range of confidence intervals.
#| fig-width:   8.5
#| fig-height:  4.5

@@

###### facet plot
cv_plot_raw_materials_label %>%
  ggplot(aes(x = variable, y = cv_by_materials)) +
  geom_point() +
  geom_errorbar(aes(ymin = cv_low_sharma,
                    ymax = cv_high_sharma), width = .2) +
    # draw a line to show our threshold value
  geom_hline(yintercept = 25, colour = "red") +
  xlab("Law Materials") +
  ylab("Coefficient of Variation on Attributes") +
  facet_wrap( ~ label, scales = "free_y") +
  coord_cartesian(ylim = c(0,130)) +
  theme_bw() 
```

## Principal Component Analysis

PCA serves to examine correlation of attributes for determining the dominant mode of cultural transmission. @fig-pca-attributes shows positive relationship between attributes. Especially, there are strong relationships between body length and maximum length, tang width and mid width, and tang length and stem length. Body length and maximum length contribute the most to the shape of stemmed points and they are most correlated to each other. Other width and tang related attributes are close to each other but a bit far from body length and maximum length. Stem width contributes the least to the overall shape. We can assume that the body length and maximum length were prioritized when the manufacturing technique of stemmed points were transmitted.

For the temporal variation, we computed PCA for difference periods. Compared to Phase 2, the influence of tang length and stem length of Phase 3 becomes more pronounced and while stem width is less effective. However, there are no notable changes in correlations over time.

```{r}
#| label: fig-pca-attributes
#| fig-cap: PCA analysis for stemmed points, for all phases combined, then for only phase two and phase three. The direction of the arrow shows the relationship between attributes; arrows pointing in the same direction indicate variables that are highly postitively correlated. The length of the arrow indicates the strength of contribution to the stemmed point shape.
#| fig-width:   7
#| fig-height:  3

library(ggbiplot)
library(ggfortify)
### PCA analysis using ggbiplot, ggfortify pkgs

PCA <- prcomp(DF, center = TRUE, scale = TRUE)

# remove stage 1 artefact
xx_pca <-
  df_sitename %>%
  dplyr::filter(SPstage.Stage != 1) %>%
  dplyr::select(-sitename, -SPstage.Stage, -SPstage.Raw_material) %>%
  prcomp(., center = TRUE, scale = TRUE)

xx_pca1 <-
  data.frame(xx_pca$x,
             stage = as.factor( df_sitename$SPstage.Stage[df_sitename$SPstage.Stage !=1]))

## CV with site (n=X)
label_phase <-
  df_sitename %>%
  group_by(SPstage.Stage) %>%
  add_tally() %>%
  mutate(label = as.factor(paste0(SPstage.Stage,'\nN = ',n)))

count_phase <- length(which(label_phase[9]>0))
count_phase2 <- length(which(label_phase[9] == 2))
count_phase3 <- length(which(label_phase[9] == 3))

# phase 2

xx_pca2 <-
  df_sitename %>%
  dplyr::filter(SPstage.Stage == 2) %>%
  dplyr::select(-sitename, -SPstage.Stage, -SPstage.Raw_material) %>%
  prcomp(., center = TRUE, scale = TRUE)

# phase 3

xx_pca3 <-
  df_sitename %>%
  dplyr::filter(SPstage.Stage == 3) %>%
  dplyr::select(-sitename, -SPstage.Stage, -SPstage.Raw_material) %>%
  prcomp(., center = TRUE, scale = TRUE)

##### using factoextra pkg
library(factoextra) #http://www.sthda.com/english/articles/31-principal-component-methods-in-r-practical-guide/112-pca-principal-component-analysis-essentials/

# Color variables by the continuous variable
cv_all <-
fviz_pca_var(xx_pca) +
  ggtitle(paste0("All phases", " (n = ", count_phase,")")) +
  theme(legend.position = "none")

# phase 2
cv_two <-
  fviz_pca_var(xx_pca2) +
  ggtitle(paste0("Phase 2", " (n = ", count_phase2,")"))

# phase 3
cv_three <-
  fviz_pca_var(xx_pca3) +
  ggtitle(paste0("Phase 3", " (n = ", count_phase3,")"))

library(patchwork)
cv_all + cv_two + cv_three
```

>>>>

@@

```{r}
#| label: fig-plot-cv-over-time
#| fig-cap: CV values for the second and third chronological phases in the Korean Late Paleolithic period. The points represent CV values for each attribute. The vertical lines indicate the range of confidence intervals.
#| fig-width:   4.5
#| fig-height:  4

@@

```{r}
#| label: fig-plot-cv-four-assemblage
#| fig-cap: CV values of four attributes from assemblages with more than 5 stemmed points. Among the four assemblages, SYG1_2 and SYG6_2 are from Phase 2 of the Korean Late Paleolithic chronology and SYG1_3 and YS_3 are from Phase 3.
#| fig-width:   4.5
#| fig-height:  4

@@

```{r}
#| label: fig-plot-cv-by-site
#| fig-cap: CV values per site for each site has more than five stemmed points. The points represent CV values for each attribute. The vertical lines are confidence intervals.
#| fig-width:   8.5
#| fig-height:  4

@@

cv_plot_site_label %>%
  ggplot(aes(x = variable,
             y = cv_by_site)) +
  geom_point() +
  geom_errorbar(aes(ymin = cv_low_sharma,
                    ymax = cv_high_sharma),
                width = 0.2) +
  xlab("Attribute") +
  ylab("Coefficient of Variation on Attributes") +
    # draw a line to show our threshold value
  geom_hline(yintercept = 25, colour = "red") +
  facet_wrap( ~ label, scales = "free_y") +
  theme(legend.position = "none") +
  coord_cartesian(ylim = c(0,70)) +
  theme_bw() +
  theme(strip.text.x = element_text(size = 6.5))
```

@@

We examined the relationship between CV values and raw materials to test the hypothesis that the shape of stemmed points was highly dependent on raw material. We excluded raw materials that were used for less than three points. @fig-plot-cv-by-raw-material shows that quartzite, the raw material for the stemmed points at Sachang, has the highest CV values, and also has wide conference intervals. Other raw materials are generally stable in variation throughout the attributes.

```{r}
#| label: fig-plot-cv-by-raw-material
#| fig-cap: CV values for raw materials that were used for making more than one stemmed point. The points represent CV values for each attribute. The vertical lines indicate the range of confidence intervals.
#| fig-width:   8.5
#| fig-height:  4

@@

###### facet plot
cv_plot_raw_materials_label %>%
  ggplot(aes(x = variable, y = cv_by_materials)) +
  geom_point() +
  geom_errorbar(aes(ymin = cv_low_sharma,
                    ymax = cv_high_sharma), width = .2) +
    # draw a line to show our threshold value
  geom_hline(yintercept = 25, colour = "red") +
  xlab("Law Materials") +
  ylab("Coefficient of Variation on Attributes") +
  facet_wrap( ~ label, scales = "free_y") +
  coord_cartesian(ylim = c(0,250)) +
  theme_bw() 
```

## Correlation analysis of attributes

```{r}
#| label: fig-corr-analysis
#| fig-cap: Correlation analysis of point attributes, calculated using Pear
#| fig-width:   7
#| fig-height:  4

library(corrr)

# correlation of all artefacts
df_sitename_corr <- 
  df_sitename %>%
  dplyr::select(-sitename, -SPstage.Stage, -SPstage.Raw_material) %>% 
  correlate() %>% 
  shave()

all_corr_plot <- rplot(df_sitename_corr, legend = FALSE)

df_sitename_corr_long <- 
  df_sitename_corr %>% 
    pivot_longer(-term)

# correlation of phase 2 artefacts
df_sitename_corr_p2 <- 
  df_sitename %>%
  dplyr::filter(SPstage.Stage == 2) %>%
  dplyr::select(-sitename, -SPstage.Stage, -SPstage.Raw_material) %>% 
  correlate() %>% 
  shave()

p2_corr_plot <- rplot(df_sitename_corr_p2, legend = FALSE)

df_sitename_corr_p2_long <- 
  df_sitename_corr_p2 %>% 
    pivot_longer(-term)

# correlation of phase 3 artefacts
df_sitename_corr_p3 <- 
  df_sitename %>%
  dplyr::filter(SPstage.Stage == 3) %>%
  dplyr::select(-sitename, -SPstage.Stage, -SPstage.Raw_material) %>% 
  correlate() %>% 
  shave()

p3_corr_plot <-  
  rplot(df_sitename_corr_p3) +
   theme(legend.position = c(0.8, 0.8))

df_sitename_corr_p3_long <- 
  df_sitename_corr_p3 %>% 
    pivot_longer(-term)

# compare p2 and p3 correlations
compare_p2_and_p3_correlations_plot <- 
list(`phase 2` = df_sitename_corr_p2_long,
     `phase 3` = df_sitename_corr_p3_long) %>% 
  bind_rows(.id = "phase") %>% 
  ggplot() +
  aes(phase,
      value) +
  geom_boxplot() +
  ggbeeswarm::geom_quasirandom(size = 4,
                               alpha = 0.3) +
  theme_minimal() +
  labs(y = "Correlation coefficient")

library(cowplot)
plot_grid(
          plot_grid(p2_corr_plot, 
                    p3_corr_plot,
                    ncol = 1,
                    labels = c("A", "B")),
          compare_p2_and_p3_correlations_plot,
          ncol = 2,
          labels = c("", "C"),
          rel_heights = c(1, 1.5))
